---
title: "Theophylline Single-Dose PK Analysis"
author: "Dylan M. Glatt, PhD"
date: "2026-05-11"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
    fig_width: 8
    fig_height: 5
  
---

```{r setup, include=FALSE}

library(tidyverse)
library(kableExtra)

# include=FALSE: this chunk runs but nothing is shown in the output, no code and no output
# echo=FALSE: this chunk runs and OUTPUT is shown, but the code is hidden 
knitr::opts_chunk$set(
  echo = TRUE, # show code by default (overridden per chunk)
  warning = FALSE, # suppress warnings
  message = FALSE, # suppress package loading messages
  dpi = 150
)

pk <- as_tibble(Theoph) # temporary add

```

## 1. Objective

This report summarizes the pharmacokinetics of theophylline following 
a single oral dose in 12 healthy adult subjects. Data are from the 
`Theoph` dataset (Boeckmann et al., 1994).

**Key parameters reported:** C~max~, T~max~, AUC~last~, and terminal half-life (t~½~).

---

## 2. Data summary

```{r data-load}
pk <- pk |>
  mutate(
    nom_time = case_when(
      Time == 0      ~ 0,
      Time < 0.4     ~ 0.25,
      Time < 0.75    ~ 0.5,
      Time < 1.5     ~ 1,
      Time < 3       ~ 2,
      Time < 4.5     ~ 4,
      Time < 6       ~ 5,
      Time < 8       ~ 7,
      Time < 10.5    ~ 9,
      Time < 18      ~ 12,
      TRUE           ~ 24
    )
  )
```

The dataset contains **`r nrow(pk)`** observations from 
**`r n_distinct(pk$Subject)`** subjects. Doses ranged from 
**`r round(min(pk$Dose * pk$Wt), 1)`** to 
**`r round(max(pk$Dose * pk$Wt), 1)`** mg.

---

## 3. Subject demographics

```{r demographics-table}
pk |>
  group_by(Subject) |>
  summarize(
    Weight_kg  = first(Wt),
    Dose_mgkg  = first(Dose),
    Total_dose = round(first(Dose) * first(Wt), 1),
    .groups    = "drop"
  ) |>
  rename(
    "Subject"        = Subject,
    "Weight (kg)"    = Weight_kg,
    "Dose (mg/kg)"   = Dose_mgkg,
    "Total Dose (mg)"= Total_dose
  ) |>
  kable(caption = "Table 1. Subject demographics and dose information") |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    font_size         = 13
  )
```

---

## 4. Concentration-time profiles

### 4.1 Individual profiles

```{r individual-profiles, fig.cap="Figure 1. Individual theophylline concentration-time profiles (semi-log scale)."}
ggplot(pk, aes(x = Time, y = conc, group = Subject, color = Subject)) +
  geom_line(linewidth = 0.6, alpha = 0.8) +
  geom_point(size = 1.8, alpha = 0.8) +
  scale_y_log10(
    breaks = c(0.5, 1, 2, 5, 10),
    labels = c("0.5", "1", "2", "5", "10")
  ) +
  annotation_logticks(sides = "l") +
  labs(
    x     = "Time after dose (h)",
    y     = "Theophylline concentration (mg/L)",
    color = "Subject"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())
```

### 4.2 Mean profile

```{r mean-profile, fig.cap="Figure 2. Mean ± SD theophylline concentration-time profile (semi-log scale). Individual data shown in gray; mean profile at nominal time shown in blue."}
pk_mean <- pk |>
  group_by(nom_time) |>
  summarize(
    mean_conc = mean(conc),
    sd_conc   = sd(conc),
    .groups   = "drop"
  )

ggplot() +
  geom_line(data = pk,
            aes(x = Time, y = conc, group = Subject),
            color = "gray70", linewidth = 0.3, alpha = 0.6) +
  geom_line(data = pk_mean,
            aes(x = nom_time, y = mean_conc),
            color = "#185FA5", linewidth = 1.2) +
  geom_point(data = pk_mean,
             aes(x = nom_time, y = mean_conc),
             color = "#185FA5", size = 2.5) +
  geom_errorbar(data = pk_mean,
                aes(x    = nom_time,
                    ymin = pmax(mean_conc - sd_conc, 0.01),
                    ymax = mean_conc + sd_conc),
                color = "#185FA5", width = 0.3) +
  scale_y_log10(breaks = c(0.5,1,2,5,10),
                labels = c("0.5","1","2","5","10")) +
  annotation_logticks(sides = "l") +
  labs(
    x       = "Time after dose (h)",
    y       = "Theophylline concentration (mg/L)",
    caption = "Blue: mean ± SD at nominal time  |  Gray: individual profiles"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())
```

---

## 5. Descriptive PK summary

```{r pk-summary-table}
# Compute simple NCA metrics manually
# Full PKNCA NCA comes in Module 9 — this is a preview
pk_params <- pk |>
  group_by(Subject) |>
  summarize(
    Dose_mg  = round(first(Dose) * first(Wt), 1),
    Cmax     = round(max(conc), 2),
    Tmax     = Time[which.max(conc)],
    Clast    = round(conc[which(Time == max(Time))], 3),
    Tlast    = max(Time),
    .groups  = "drop"
  )

pk_params |>
  kable(
    caption = "Table 2. Descriptive PK parameters by subject",
    col.names = c("Subject","Dose (mg)","Cmax (mg/L)",
                  "Tmax (h)","Clast (mg/L)","Tlast (h)")
  ) |>
  kable_styling(
    bootstrap_options = c("striped","hover","condensed"),
    full_width        = FALSE,
    font_size         = 13
  ) |>
  column_spec(3, bold = TRUE, color = "#185FA5")
```

### Summary statistics

```{r summary-stats}
pk_params |>
  summarize(
    `Mean Cmax (mg/L)`  = round(mean(Cmax), 2),
    `SD Cmax`           = round(sd(Cmax), 2),
    `CV% Cmax`          = round(sd(Cmax)/mean(Cmax)*100, 1),
    `Median Tmax (h)`   = median(Tmax),
    `Range Tmax (h)`    = paste0(min(Tmax), "–", max(Tmax))
  ) |>
  kable(caption = "Table 3. Summary of Cmax and Tmax across subjects") |>
  kable_styling(
    bootstrap_options = c("striped","condensed"),
    full_width        = FALSE,
    font_size         = 13
  )
```

---

## 6. Conclusions

- Theophylline reached peak concentrations (C~max~) of 
  `r round(mean(pk_params$Cmax), 2)` ± 
  `r round(sd(pk_params$Cmax), 2)` mg/L (mean ± SD) at a median 
  T~max~ of `r median(pk_params$Tmax)` hours.
- The semi-log profiles show apparent log-linear decline after T~max~, 
  consistent with first-order elimination kinetics.
- Full NCA parameter estimation (AUC, t~½~, CL/F) will be performed 
  in Module 9 using PKNCA.

---

*Report generated: `r format(Sys.time(), "%Y-%m-%d %H:%M")`*  
*R version: `r R.version$version.string`*

